---
title: "Study population Khayelitsha"
author: "Remo Schmutz"
date: "2023-11-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)
library(table1)
library(tidyverse)
library(knitr)
library(png)
library(janitor)
library(flextable)
library(forcats)
```

## Study population

I defined ART-naive patients as scoring 1 or NA in naive_y and simultaneously scoring 1 or NA in naive_y. Also i filtered out patients with a enrollment data more than a month after their ART start (encart_d). This is the code i use to determine ART-naive patients:

```{r}
tblBAS <- read.csv("../../data_raw/RSA/tblBAS_incl_naive_imp2.csv") %>% 
  mutate(across(c(birth_d, enrol_d, recart_d), ~na_if(.x, "")),
         across(c(birth_d, enrol_d, recart_d), ~as.Date(.x, format = "%Y-%m-%d")),
         age_at_art_start = interval(birth_d, recart_d) %/% years(1),
         sex = as.factor(case_when(sex == 1 ~ "Male",
                         sex == 2 ~ "Female",
                         TRUE ~ NA))) %>%  
  rename(born = birth_d, 
         art_start_date = recart_d) %>% 
  distinct(patient, .keep_all = TRUE) %>% 
  select(patient, enrol_d, born, sex, art_start_date, age_at_art_start, naive_y, naive_y_imp)

tblNAIVE <- tblBAS %>% 
  filter(naive_y_imp %in% c(1,NA) & naive_y %in% c(1,NA)) %>% 
  mutate(diff = enrol_d - art_start_date) %>% 
  filter(diff <= 30 | is.na(diff))
```


This leaves us with ``r nrow(tblNAIVE)`` patients.

Using our exclusion/inclusion criteria: 
```{r}
tblPOP <- tblNAIVE %>% 
  filter(between(art_start_date, as.Date("2010-01-01"), as.Date("2022-12-31")),
         !is.na(sex),
         !is.na(born) & !is.na(art_start_date),
         age_at_art_start >= 16)
```

This leaves us with ``r nrow(tblPOP)`` patients.

## WHO stage

To calculate the baseline WHO-stage, I use the tblVIS dataset and check visits 180 days before and after ART start. For every patient I use the closest non-NA entry to determine his WHO-stage at ART start. Using such a wide time-interval is a rather generous approach. The results are displayed in Table 1.

```{r}
tblVIS <- read.csv("../../data_raw/RSA/tblVIS.csv") %>% 
  filter(patient %in% tblNAIVE$patient) %>% 
  select(patient, vis_d, who_stage) %>% 
  mutate(vis_d = as.Date(vis_d, format = "%Y-%m-%d")) %>% 
  left_join(tblBAS %>% select(patient, art_start_date), by = "patient") %>% 
  group_by(patient) %>%
  mutate(difference = abs(vis_d - art_start_date),
         who_stage = as.factor(case_when(who_stage %in% c(2,3) ~ "2/3",
                               who_stage == 9 ~ NA,
                               TRUE ~ as.character(who_stage)))) %>%
  filter(difference <= 180) %>%
  arrange(patient, is.na(who_stage), difference) %>% 
  slice(1) %>%
  ungroup() %>%
  select(patient, who_stage)
```


## Table 1

```{r, include=FALSE}

complete <-  readRDS("../../data_clean/art.rds")
ch <- readRDS("../../data_clean/ch/art_ch.rds")
rsa <- readRDS("../../data_clean/rsa/art_rsa.rds")

# Columns #

complete <- complete %>%
  mutate(presenting_tb = factor(presenting_tb, levels = c(0, 1))) %>%
  mutate(presenting_tb = 
           fct_recode(presenting_tb,
                            "Not presenting with TB" = "0",
                            "Presenting with TB" = "1"),
         site_tb = case_when(presenting_tb == 0 ~ NA,
                             TRUE ~ site_tb))

# Age #

label(complete$age_at_art_start) <- "Age at start of ART"

# Sex #

complete$gender <- factor(complete$gender, 
         levels=c("Male","Female"))
label(complete$gender) <- "Gender"

# Site of TB #

complete <- complete %>% 
  mutate(site_tb = case_when(presenting_tb == "Not presenting with TB" ~ NA,
                             TRUE ~ site_tb),
         regimen_tb_group = case_when(
           presenting_tb == "Not presenting with TB" ~ NA,
           TRUE ~ regimen_tb_group))
label(complete$site_tb) <- "Site of TB"

# Source of original HIV infection #

# complete$risk <- 
#   factor(complete$risk, 
#          levels=c("Homosexual","Heterosexual","IVD, needle sharing", "IVD/sexual", 
#                   "Other blood products", "Perinatal", "Other sources"),
#          labels = c("Homosexual contacts", "Heterosexual contacts", "I.v. drug use (with needle sharing)", "I.v. drugs/sexual contacts (unclear which one)", "Other blood products", "Perinatal transmission", "Other sources"))
# 
# label(complete$risk) <- "Source of original HIV infection"

# CD4 baseline #

label(complete$cd4_baseline) <- "CD4 cell count at start of ART"

# Viral load baseline #

label(complete$rna_baseline) <- "HIV RNA viral load at start of ART"

# WHO stage #

label(complete$who_stage) <- "WHO clinical stage"

# HIV treatment regimens #

label(complete$regimen) <- "HIV treatment regimens"

# TB treatment regimens #

label(complete$regimen_tb_group) <- "TB regimen"


```

```{r echo=FALSE, fig.align='center'}
my.render.cont <- function(x) {
    stats <- stats.default(x)
    with(stats, {
        mean_sd <- sprintf("%d (%d)", round(MEAN), round(SD))
        median_iqr <- sprintf("%d [%d, %d]", round(MEDIAN), round(Q1), round(Q3))
        min_max <- sprintf("[%d, %d]", round(MIN), round(MAX))
        c("", "Mean (SD)"=mean_sd, "Median (IQR)"=median_iqr, "Min, Max"=min_max)
    })
}

my.render.cat <- function(x) {
    # Calculate the total number of non-missing values
    non_missing_total <- sum(!is.na(x))
    
    # Use stats.default to compute frequencies and percentages
    stats <- stats.default(x)
    
    # Use sapply to apply the function to each element of the stats list
    c("", sapply(stats, function(y) {
        # Calculate the percentage excluding missings
        real_pct <- if (non_missing_total > 0) {
            (y$FREQ / non_missing_total) * 100
        } else {
          NA  # Avoid division by zero if non_missing_total is zero
        }
        
        # Prepare the string with both percentages: including and excluding missings
        sprintf("%d (%0.0f%%; %0.0f%%)", y$FREQ, y$PCT, real_pct)
    }))
}


table1(~ age_at_art_start + gender + site_tb + cd4_baseline + rna_baseline + who_stage + regimen + regimen_tb_group| presenting_tb, data = complete %>% filter(cohort == "RSA"),
       overall = c(left = "All"),
       topclass="Rtable1-zebra",
       caption = "Baseline characteristics of HIV-positive patients starting antiretroviral treatment (ART) between 2010 and 2022, presenting and not presenting with TB. Patients with a TB diagnosis >12 months before ART start were excluded. We compared patients with prevalent TB (all forms, diagnosis 2 months before/after ART start) or a recent history of TB (diagnosis 2-12 months before ART start) with patients with no history of TB.",
       render.continuous = my.render.cont,
       render.categorical = my.render.cat)
```
